Type Error Telemetry at Scale

### 2023-09-28 programming23/paper.tex
sec 5

With this telemetry data, we investigate research questions about
the adoption and benefits of type analysis:
\begin{description}
  \item[RQ1.]
    How many creators use type analysis?
    How often do projects contain modules with different
    analysis modes?
    How often do creators turn analysis off?
  \item[RQ2.]
    For modules that use type analysis:
    which errors arise,
    how do creators respond,
    and which errors tend to persist despite subsequent edits?
  \item[RQ3.]
    What impact does type analysis have on the number of \FS{}
    errors?
    For example, do \FS{} errors pile up in unanalyzed projects?
\end{description}

@@

With the data in hand, we can answer the research questions
from~\cref{s:introduction}.
The answers confirm several aspects of Luau's design, but
at the same time offer several directions for future improvements.
Luau clearly has a powerful typechecker that can efficiently
analyze thousands of lines of code.
But, few creators use type analysis directly and the data presents little
evidence that doing so would improve their quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

@@

  \item[How often do projects contain modules with different analysis modes?]
    Projects rarely combine analysis modes.
    Only 512 of \nK{340} sessions ($<\pct{1}$) switched to modules with different
    analysis modes.
    This number may be a significant underapproximation because we have data
    only on modules that the creator chose to open, but the widespread use
    of \mnocheck{} mode suggests that it is not off by much.

  \item[How often do creators turn analysis off?]
    Creators rarely disable type analysis after opting in.
    Only 233 of the \nK{340} sessions ($<\pct{1}$)
    switch to a weaker mode.
    Most of these (\pct{75}) are downgrading from \mstrict{}
    rather than \mnonstrict{} mode.
    %% FILL more insight if we match upgrades + downgrades in session?

@@

  \item[For modules that use type analysis:]\leavevmode
    \begin{description}
      \item[- Which errors arise?]
        Syntax errors are extremely common~(over \pct{50},~\cref{t:type-error-count}), 
        followed by type mismatches~(\pct{20} in \mstrict{}),
        arity errors~(\pct{2}),
        and failure to unpack on optional value~(\pct{2}).
        There are several uncommon errors, such as misusing a table operation,
        and a few that never arise: internal and too-complex errors,
        duplicating or swapping a generic parameter, use of a deprecated API,
        and unsafe dynamic property lookup on a class.

      \item[- How do creators respond? Which errors persist through edits?]
        Creators typically fix analysis errors.
        With few exceptions, increases and decreases in the number of errors
        balance out~(\cref{f:error-changes,t:type-error-survival,}).
        Even in \mnocheck{}, most edits that overlap with syntax errors
        remove at least one error~(\pct{67}) rather than increase or maintain
        the total.
        The errors that most-often survive edits are due to
        inference at a binary operation,
        sending \code{self} to a function that does not expect it,
        and failure to unpack an optional value.
        But the first two seldom arise, and the third is something creators
        might reasonably ignore while prototyping.

@@

Third, the total counts of type errors include a large number of syntax errors,
which have little to do with the Luau type system and are presumably easy for
creators to fix.
It would be better for the analysis to ignore these errors.
Indeed, our comparison between \FS{} errors in \mnocheck{} vs.
\mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
without all the noise that syntax errors introduce.

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Although it appears that creators choose to fix errors with their edits,
the fixes may be a side effect of other plans.

@@

Lastly, we have two remarks about external validity.
First, our information on type errors is limited to error names.
If another language has an error that is roughly similar to the
\code{CannotCallNonFunction} error but with slightly different
semantics, then the relative frequency of that error could be different
than what we observed.
Second, we know nothing about the creators in our study except that they
were selected at random from a large and diverse group.
A dataset based on a targeted subset of users might show entirely different
characteristics.

>>>>

With this telemetry data, we investigate research questions about
the adoption and benefits of type analysis: \todo{Some of these are flawed. We don't know anything about creators. Adjust
  to match section 5.}
\begin{description}
  \item[RQ1.]
    How many creators use type analysis?
    How often do projects contain modules with different
    analysis modes?
    How often do creators turn analysis off?
  \item[RQ2.]
    For modules that use type analysis:
    which errors arise,
    how do creators respond,
    and which errors tend to persist despite subsequent edits?
  \item[RQ3.]
    What impact does type analysis have on the number of \FS{}
    errors?
    For example, do \FS{} errors pile up in unanalyzed projects?
\end{description}

@@

\todo{No identity $\Rightarrow$ no ``creators'', only sessions.}
With the data in hand, we can answer the research questions
from~\cref{s:introduction}.
The answers confirm several aspects of Luau's design, but
at the same time offer several directions for future improvements.

Luau clearly\todo{How can we say?} has a powerful typechecker that can efficiently
analyze thousands of lines of code.
But, few creators use type analysis directly and the data presents little
evidence that doing so would improve their quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

@@

  \item[How often do projects contain modules with different analysis modes?]
    Projects rarely combine analysis modes.
    Only 512 of \nK{340} sessions ($<\pct{0.15}$) switched to modules with different
    analysis modes.
    This number may be a significant underapproximation because we have data
    only on modules that the creator chose to open, but the widespread use
    of \mnocheck{} mode suggests that it is not off by much.

  \item[How often do creators turn analysis off?]
    Creators rarely disable type analysis after opting in.
    Only 233 of the \nK{340} sessions ($<\pct{0.07}$)
    switch to a weaker mode.
    Most of these (\pct{75}) downgrade from \mstrict{} to
    \mnonstrict{} % BEN: confirm!
    rather than from \mnonstrict{} mode.
    %% FILL more insight if we match upgrades + downgrades in session?

@@

  \item[For modules that use type analysis (\mstrict{} and \mnonstrict{}):]\leavevmode
    \begin{description}
      \item[- Which errors arise?]
        Syntax errors are extremely common~(over \pct{50},~\cref{t:type-error-count}), 
        followed by type mismatches~(\pct{20} in \mstrict{}),
        arity errors~(\pct{2}),
        and failure to unpack on optional value~(\pct{2}).
        There are several uncommon errors, such as misusing a table operation,
        and a few that never arise: internal and too-complex errors,
        duplicating or swapping a generic parameter, use of a deprecated API,
        and unsafe dynamic property lookup on a class.

      \item[- How do creators respond? Which errors persist through edits?]
        Creators typically fix analysis errors.
        With few exceptions, increases and decreases in the number of errors
        balance out~(\cref{f:error-changes,t:type-error-survival,}).
        Even in \mnocheck{}, most edits that overlap with syntax errors
        remove at least one error~(\pct{67}) rather than increase or maintain
        the total.
        The errors that most often survive edits are due to
        inference at a binary operation,
        sending \code{self} to a function that does not expect it,
        and failure to unpack an optional value.
        But the first two seldom arise, and the third is something creators
        might reasonably ignore while prototyping.

@@

Third, the total counts of type errors include a large number of syntax errors,
which have little to do with the Luau type system and are presumably easy for
creators to fix.
It would be better for the analysis to ignore these errors.
Indeed, our comparison between \FS{} errors in \mnocheck{} vs.
\mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
without all the noise that syntax errors introduce.\todo{Why didn't
  YOU just filter them, then?!?}

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Thus, fixes may be a side effect of other plans, as seen in prior work [CITE].
% Cite the Marceau SIGCSE 2011 paper. This is the [DIFF] code from the rubric.

@@

Lastly, we have two remarks about external validity.
First, our information on type errors is limited to error names.
If another language has an error that is roughly similar to
the\todo{Huh? It's similar yet different?}
\code{CannotCallNonFunction} error but with slightly different
semantics, then the relative frequency of that error could be different
than what we observed.
Second, we know nothing about the creators in our study except that they
were selected at random from a large and diverse group.
A dataset based on a targeted subset of users might show entirely different
characteristics.

Ultimately, we view our work as primarily \emph{formative}. It gives
us a first look at usage, and generates several questions that can
lead to hypotheses. Answering those would require much more intrusive
techniques, or controlled studies, or other means.
